---
title: "The Dashboard"
subtitle: "Bench the Instinct — NBA Player Risk Tool for Coaches & GMs"
description: "An interactive load management dashboard predicting which NBA players are at risk of missing significant game time"
toc: false
page-layout: full
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

::: {.hero-banner}
# Bench the Instinct 🏀

::: {.badge-red}
INTERACTIVE TOOL
:::

<br>

*"We know load caused the spike. Now we know who's next. The question is whether
the league acts on data — or waits for the next Tatum."*

<br>

Built for **coaches, GMs, and medical staff** who need a data-backed answer to
the question they have been asking on instinct for years: *should we rest this player?*
:::

---

## What the Dashboard Does

::: {.grid}

::: {.g-col-4}
::: {.part-card}
### 🦵 The Spike
Visual context for why this tool exists. The 2024–25 rupture spike, load trends over time, and the confirmed injury record since 1990.
:::
:::

::: {.g-col-4}
::: {.part-card .red}
### 🔮 Who's Next?
Filterable risk table of 2024 NBA players ranked by predicted % of next season's games missed. Restricted to meaningful contributors (≥20 rolling minutes).
:::
:::

::: {.g-col-4}
::: {.part-card}
### ⚡ Bench the Instinct
The load simulator. Enter a player's rolling minutes, prior absence history, and career tenure. Get a predicted risk score. Compare scenarios before making the call.
:::
:::

:::

---

## Launch the Dashboard

::: {.text-center style="padding: 40px 0;"}

[🏀 Open Full Dashboard](https://7tmnpt-keziah-vickraman.shinyapps.io/nba-achilles/){.btn .btn-danger .btn-lg target="_blank" style="font-size: 1.2rem; padding: 16px 40px; margin-right: 16px;"}

[📄 Download Academic Paper](submission_final_042026.pdf){.btn .btn-primary .btn-lg target="_blank" style="font-size: 1.2rem; padding: 16px 40px;"}

:::

---

## In-Page Preview {#preview}

::: {.callout-tip}
## Tip
The embedded preview below may take 10–15 seconds to load. For the best experience
use the **Open Full Dashboard** button above.
:::

```{=html}
<div style="border: 2px solid #17408B; border-radius: 8px; overflow: hidden; margin: 20px 0;">
  <div style="background: #17408B; padding: 8px 16px; color: white; font-family: 'Roboto Condensed', sans-serif; font-weight: 700;">
    🏀 Bench the Instinct — NBA Risk Dashboard
    <span style="float:right; font-size: 0.85rem; font-weight: 400; opacity: 0.8;">
      Powered by R Shiny
    </span>
  </div>
  <iframe 
    src="https://7tmnpt-keziah-vickraman.shinyapps.io/nba-achilles/" 
    width="100%" 
    height="800px" 
    frameborder="0"
    style="display: block;">
  </iframe>
</div>
```

---

## Model at a Glance

```{r model-summary}
#| label: tbl-model-summary
#| tbl-cap: "Model performance on held-out test set (2022–2025)"

tibble::tribble(
  ~Metric,  ~Value,   ~`Plain English`,
  "R²",     "0.747",  "Explains 74.7% of variance in games missed",
  "RMSE",   "0.133",  "Average prediction error ~11 games on 82-game scale",
  "MAE",    "0.100",  "Median prediction error ~8 games on 82-game scale"
) |>
  knitr::kable(align = c("l", "r", "l"))
```

::: {.callout-important}
## Honest limitations
`pct_games_missed` captures **all** absence reasons — Achilles injuries, other injuries,
rest decisions, and suspensions. The model predicts load-related game unavailability
as a proxy for injury burden. It does not diagnose specific injury types.

`back_to_back_games` and `total_minutes` were removed from the model after PDP
analysis revealed survivorship bias — healthy players accumulate more of both,
so the features had inverted relationships with the outcome. The primary load
predictor is `avg_rolling_minutes_3` (lagged 3-game rolling average minutes),
which correctly shows a positive monotonic relationship with games missed.

Full methodology in the [academic paper](submission.qmd).
:::

---

## Who Should Use This

| Role | How to use it |
|---|---|
| **Head coach** | Run the what-if simulator before a heavy road trip — see how reduced minutes changes predicted risk |
| **GM / front office** | Filter the risk table before contract decisions — who is likely to miss 20+ games next season? |
| **Medical / performance staff** | Flag high-risk players for targeted prehabilitation or biomechanical screening |
| **League operations** | Aggregate risk scores by team to identify scheduling interventions |

---

*Dashboard built in R Shiny · Model: XGBoost regression · Data: hoopR (2018–2025)*
*Keziah Vickraman · Capstone 2025*
