---
title: "Did Load Kill the Achilles?"
subtitle: "A Causal Investigation into Whether the League's Scheduling and Load Patterns Drove the 2024–25 Rupture Spike — and What the League Should Do About It"
description: "Difference-in-Differences analysis of NBA player workload and confirmed Achilles ruptures (1990–2025)"
date: "April,2026"
author:
  - name: Keziah Vickraman
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(scales)
library(glue)
library(jtools)
library(knitr)
load("00_Shared_Data.RData")
load("Part1_Causal_Results.RData")
```

---

*Imagine if stars like Tatum, Haliburton, and KD could have avoided injuries in
those crucial games. No more "in the perfect world", or "in another world" — no more
'what ifs.' This study aims to turn what-ifs into actionable foresight.*

---

## Introduction

Achilles tendon ruptures are among the most catastrophic musculoskeletal injuries
sustained by professional athletes, particularly in high-impact sports such as
basketball. In the National Basketball Association (NBA), where explosive movements,
rapid directional changes, and consistently high workloads are the norm, a ruptured
Achilles can significantly disrupt or even prematurely end a player's career.
Although less frequent than other lower extremity injuries, Achilles tendon ruptures
carry disproportionately severe consequences. High-profile cases such as Kobe Bryant
and Kevin Durant have drawn public attention to the profound physical, psychological,
and financial impact of this injury (Farmer, 2013; Hart, 2019; Lewis, 2014).

The average recovery time following an Achilles rupture ranges from 9 to 12 months,
yet many athletes never fully return to their prior level of performance. Studies have
shown that even when return to play is achieved, players often experience lasting
declines in minutes played, efficiency, and longevity (Amin et al., 2013; Trofa et al.,
2017). A systematic review by Trofa et al. (2017) found that although most elite
athletes return to competition, performance metrics generally decrease, especially among
older or high-usage players. These outcomes underscore the injury's severity and its
long-term implications for career sustainability.

---

## Past Research

Research has primarily focused on post-injury recovery and return-to-play timelines.
Petway et al. (2022), for instance, conducted a biomechanical analysis of NBA gameplay
footage and identified specific loading and landing patterns commonly preceding Achilles
tendon ruptures, implying that such injuries may not be entirely random. Likewise,
LaPrade et al. (2022) emphasized the role of cumulative fatigue and overuse —
especially during congested playing schedules — as likely contributors to tendon
failure in elite athletes.

Despite these insights, there remains a lack of robust, data-driven tools to
proactively identify players at elevated risk of Achilles tendon rupture. Given the
NBA's intense physical demands and the substantial financial and competitive costs
associated with player injuries, the development of predictive models is both timely
and essential. As Siu et al. (2020) argued, early identification of at-risk players
can support preventative strategies such as load management, individualised rest
protocols, and targeted prehabilitation — which may ultimately reduce injury incidence
and prolong athletic careers.

More recently, Achilles tendon injuries appear to be increasing in visibility and
possibly in frequency. Between 1990 and 2023, 45 NBA players sustained Achilles tendon
ruptures — an average of 1.36 cases per season (Goodwill, 2024). Alarmingly, three
such injuries occurred during the 2023 postseason alone, prompting renewed concerns
about underlying risk factors (Fischer, 2023). In response, NBA Commissioner Adam
Silver acknowledged that the league is actively researching the causes of these
injuries, including the intensifying pace of the game, tighter travel schedules, and
player workload demands (NBA.com, 2024). While surgical and rehabilitative advances
have improved post-injury outcomes (LaPrade et al., 2022; Trofa et al., 2017), the
question of why certain athletes rupture their Achilles in the first place remains
less understood.

---

## The 2024–25 Spike — Is This Just Me, or Did It Spike?

The 2024–25 season changed the conversation entirely.

```{r spike-plot, fig.height=5, fig.cap="Confirmed NBA Achilles ruptures by year (1990–2025). Orange highlight = 2024–25 season. Dashed line = historical average of 1.96 ruptures per year."}
spike_VIZ
```

Eight confirmed Achilles ruptures in a single season. Tyrese Haliburton. Jayson Tatum.
Damian Lillard. Dejounte Murray. The names read like an All-Star ballot, not an injury
report. The 2024–25 season recorded nearly **6× the historical per-season average** of
1.36 ruptures. This is not random variation — this is a structural signal.

NBA Commissioner Adam Silver convened an expert panel in response, stating the league
is taking a serious look at Achilles tears and pointing specifically to workload
intensity and scheduling density as primary investigative targets (Silver, 2025;
Reynolds, 2025). This study is our analytical response to that question.
```{r}
#| label: rupture-curation
#| echo: true
#| code-fold: true
#| code-summary: "Show data curation code"

achilles_ruptures_FULL %>%
  select(player_name, injury_date, injury_type, injury_severity) %>%
  mutate(injury_date = format(as.Date(injury_date), "%B %Y")) %>%
  arrange(injury_date) %>%
  rename(Player = player_name,
         Date = injury_date,
         Type = injury_type,
         Severity = injury_severity) %>%
  knitr::kable()
```

::: {.callout-note}
## Data note
We identified 53 unique NBA players who suffered a confirmed Achilles tendon rupture
between 1990 and 2025. This list was curated by filtering historical NBA injury data
for confirmed rupture keywords (e.g., *tear*, *surgery*, *repair*) and validated
against public records including Fadeaway World's documented timeline (Goodwill, 2024).
The 2024–25 cases (n = 8) were manually curated from verified ESPN and Yahoo Sports
reports due to their recency.
:::

---

## What Has Changed — The Load Landscape

Before testing causality, we first ask: has player load actually changed? If the
answer is no, the DiD has nothing to identify.

```{r load-trend, fig.height=4.5, fig.cap="Rolling 3-game average minutes over time. Median load (blue) has risen post-2017 while elite player load (75th percentile, red dashed) has plateaued. The interquartile range (blue band) is widening. Dashed vertical line = 2017–18 schedule reform."}
load_trend_VIZ
```

The chart reveals a more nuanced story than a simple claim that load went up.
The median player (blue line) has been playing *more* minutes post-2017, rising from
approximately 19–20 to 21–22 minutes rolling average. The 75th percentile (red dashed),
representing high-usage starters and stars, has slightly *decreased* and plateaued.
The gap between them is widening.

This suggests **load management has protected the top end of the distribution while
average player workloads have increased** — creating a growing divergence. Players with
high usage who fall just outside formal load management protocols may be the most
structurally exposed group. This is the population the DiD analysis targets.

---

## Study Design — The Causal Framework

### What Is Difference-in-Differences?

A Difference-in-Differences (DiD) estimator compares the *change* in outcomes for a
treated group relative to a control group, before and after a structural shift. Here:

- **Treated group**: player-seasons in the top 25th percentile of rolling 3-game
  average minutes — the high-load players
- **Control group**: player-seasons below that threshold — the low-load players
- **Post period**: 2017 onwards — the 2017–18 NBA schedule reform, the most
  comprehensive structural change to player scheduling in the league's modern era
- **Outcome**: confirmed Achilles rupture (binary, per player-season)

The DiD estimator is the coefficient on the interaction term `High Load × Post-2017`.
A positive coefficient means high-load players saw a differential increase in rupture
rates after 2017, over and above any general time trend.

### Why 2017?

The 2017–18 season is the cleanest structural break in NBA scheduling history. It was
the first season to simultaneously eliminate: (a) four-games-in-five-nights stretches,
which had occurred more than 70 times per season league-wide; (b) 18-games-in-30-days
windows; and (c) single road games requiring more than 2,000 miles of travel. The NBA
also added a full week to the regular-season calendar that year (Windhorst, ESPN, 2017).
By 2018–19, no team played more than 15 back-to-back sets — down from a league average
of nearly 20 just three seasons earlier (Sportico, 2022).

We also define a **secondary cutpoint at 2023** for the Player Participation Policy
(PPP), which mandated that All-Star-level players appear in national television games
and penalised teams for resting stars. This second cutpoint is used in the selective
protection analysis below to test whether the PPP partially reversed the protective
effects of the 2017 reform.

### The B2B Moderator

Commissioner Silver specifically flagged back-to-back scheduling as a contributing
factor. We include a triple interaction — **Load × Post × B2B** — to test whether
the load effect is amplified for players with heavy B2B exposure.

### What We Investigate

This study investigates three core questions:

1. Whether players who accumulate longer playing durations over three consecutive games
   exhibit a higher risk of injury
2. Whether insufficient rest between games — especially in back-to-back contexts —
   contributes to greater rupture incidence
3. Whether the post-2017 schedule reform era has differentially affected rupture rates
   for high-load versus low-load players, and whether this protective effect varied
   by player tier (stars vs. reserve players)

Understanding these predictors can empower team managers and coaching staff to make
data-driven decisions about player workload and recovery. By accurately forecasting
injury risk, this analysis can guide strategies such as optimising rest periods and
managing player minutes — ultimately contributing to maintaining player health,
prolonging careers, and enhancing overall team performance throughout the demanding
NBA season (Siu et al., 2020).

---

## Parallel Trends Check

The validity of DiD rests on the **parallel trends assumption**: that absent the
treatment, the two groups would have trended similarly over time.

```{r parallel-trends, fig.height=5.5, fig.cap="Parallel trends check. Pre-2017 trends are broadly parallel between load groups, particularly in the Low B2B panel. Post-2017 the high-load gap widens. Dashed line = 2017–18 schedule reform. Dotted line = 2023–24 PPP."}
did_parallel_trends_VIZ
```

::: {.callout-tip}
## Reading this chart
If the blue (low load) and red (high load) lines move roughly in parallel
**before** the dashed 2017 line, the parallel trends assumption holds.
A divergence **after** 2017 that is larger for the high-load group is the causal signal.
The dotted 2023 line marks the PPP — watch whether the high-load line shifts again
after that point. Year-to-year volatility reflects the rarity of the outcome — with a
base rupture rate of ~0.29%, a single rupture in a season-load group moves the mean
materially.
:::

---

## DiD Results

::: {.panel-tabset}

### Regression Table

```{r regression-table, results='asis'}
#| code-fold: true
#| code-summary: "Show DiD model results"
jtools::export_summs(
  model_did_1, model_did_2, model_did_3,
  model.names  = c("Model 1\nLoad × Post",
                   "Model 2\nLoad × Post × B2B",
                   "Model 3\nContinuous"),
  error_format = "CIs: [{conf.low}, {conf.high}]",
  digits       = 4
)
```

### Coefficient Plot

```{r coef-plot, fig.height=4.5, fig.cap="DiD interaction coefficients across all three models. Error bars = 95% CI. All interaction terms are positive and directionally consistent with the load hypothesis."}
did_coef_VIZ
```

:::

The interaction coefficient `High Load × Post-2017` is **positive** across all three
models (Model 1: β = `r round(bayesian_stats$beta_interact, 4)`). Interpreting this
through a Bayesian lens yields a concrete risk update.

Starting from a **prior rupture probability of `r bayesian_stats$prior`%** — the
historical base rate across all player-seasons — a high-load player-season in the
post-2017 era carries an estimated **posterior rupture probability of
`r bayesian_stats$posterior`%**. This represents a **`r bayesian_stats$relative_risk`×
relative risk** over baseline, and an absolute increase of
`r bayesian_stats$abs_increase` percentage points.

The approximate **Bayes Factor is `r bayesian_stats$bayes_factor`**, meaning that
observing high load in the post-reform era is approximately twice as probable among
player-seasons that resulted in rupture than among those that did not.

::: {.callout-note}
## Interpreting a Bayes Factor of ~2 at low prevalence
By Jeffreys' scale, a Bayes Factor of 1–3 constitutes weak-to-anecdotal evidence in
isolation. However, this framing understates the practical significance when the prior
is as low as 0.29%. When the base rate is this rare, *any* evidence that doubles the
posterior odds is clinically meaningful. An Achilles rupture costs a player 9–12 months
of recovery, frequently causes permanent performance decline, and carries a mean
economic cost of $4M per player (Meadows et al., 2024). In this decision context, a
2× upward revision in risk is not a null result — it is an actionable signal that
justifies proactive load intervention.
:::

Model 2 shows further amplification when high load is combined with heavy B2B exposure
— directly corroborating Commissioner Silver's stated concern. Confidence intervals
are wide throughout, a direct consequence of the low base rate, and all results are
interpreted directionally rather than as precise effect size estimates.

---

## Who Did the Reform Actually Protect?

The DiD results show a population-level effect. But there is a more precise question
buried underneath: **did the 2017 schedule reform protect all players equally, or did
it disproportionately shelter star-tier players while leaving reserve players
structurally exposed?**

This is the **selective protection hypothesis** — and the 2023 PPP gives us a way to
test its reversal.

### Why Stars and Reserves May Respond Differently

Both the 2017 reform and the 2023 PPP were framed explicitly around high-profile
players. Teams proactively managed the minutes of their most valuable assets; rotation
players and reserves continued to absorb heavy schedules regardless. The NBA's own
PPP defines eligibility by All-Star and All-NBA status — it is, by design, a
star-focused instrument. When the PPP then mandated stars play in national TV games,
it effectively reversed the load protection that had been informally extended to them
since 2017.

We define three player tiers by season-level scoring percentile — a defensible proxy
for the NBA's own star classification:

| Tier | Definition |
|------|------------|
| **Star** | Top 15% of scoring seasons — All-Star / All-NBA calibre |
| **Starter** | 50th–84th percentile |
| **Reserve** | Bottom 50% |

### Tier Trends Across Three Policy Eras

```{r tier-trends, fig.height=5, fig.cap="Mean Achilles rupture rate by player tier across three policy eras. If the Star line drops post-2017 but Reserve stays flat, selective protection is confirmed. If Star rises post-2023, the PPP reversed that protection."}
did_tier_VIZ
```

::: {.callout-tip}
## What to look for
- **Star line drops post-2017, Reserve stays flat** → the reform disproportionately
  protected high-profile players. Reserve players remained structurally exposed.
- **Star line rises post-2023** → the PPP forced stars back into high-load game
  situations, reversing the protection. This is the mechanistic explanation for
  why the 2024–25 spike hit names like Tatum, Haliburton, Lillard and Murray —
  precisely the players whose load had previously been managed most carefully.
:::

### Selective Protection Model (Model 4)

::: {.panel-tabset}

### Regression Table

```{r tier-regression, results='asis'}
#| code-fold: true
#| code-summary: "Show Model 4 results"
jtools::export_summs(
  model_did_1, model_did_tier,
  model.names  = c("Model 1\nBaseline Load × Post",
                   "Model 4\nTier × Post\n(Selective Protection)"),
  error_format = "CIs: [{conf.low}, {conf.high}]",
  digits       = 4
)
```

### Coefficient Plot

```{r tier-coef-plot, fig.height=4.5, fig.cap="Model 4 coefficients. The 'Star × Post-2017' term is the selective protection estimator — a negative value means stars saw a larger reduction in rupture rates post-reform relative to reserve players."}
did_tier_coef_VIZ
```

:::

The `Star × Post-2017` interaction term is the selective protection estimator. A
negative coefficient means star-tier players experienced a **larger reduction** in
rupture rates after the 2017 reform relative to reserve players. A near-zero
coefficient on the reserve baseline confirms the reform offered little protective
benefit to lower-tier players.

As before, confidence intervals are wide — a direct consequence of the ~0.29% event
rate. These results are interpreted directionally, with the tier trend chart as the
primary visual evidence.

---

## Implications — How Should the League, Management and Coaches Move Forward?

The DiD analysis, together with the selective protection findings, points to four
concrete conclusions:

**1. The 2017 reform worked — but only for stars.**
The tier analysis shows star-tier rupture rates trended down after the 2017 schedule
reform, while reserve player rates stayed flat. Load management, applied informally to
high-value players, had a measurable protective effect. The problem is it never
reached the rest of the roster.

**2. Load management must reach the middle tier, not just stars.**
The most structurally exposed group is high-usage rotation players who are not famous
enough to receive formal rest protocols but are absorbing the workload that stars shed.
Load management needs to be systematic across rosters — not reserved for max-contract
names — with data-driven thresholds for all players above a minutes-per-game cutoff.

**3. The 2023 PPP may have created the 2024–25 spike.**
By mandating that All-Star players appear in national TV games, the PPP reversed the
informal load protection that had been quietly extended to stars since 2017. The tier
trend chart shows star rupture rates rising in the PPP era — and the 2024–25 casualty
list (Tatum, Haliburton, Lillard, Murray) is precisely the All-Star cohort the PPP
targeted. This is a policy unintended consequence the league needs to acknowledge and
correct.

**4. Back-to-back scheduling remains a structural risk factor.**
The triple interaction shows the load effect is amplified in B2B contexts. While the
NBA has reduced B2B games since 2017, the remaining B2B games disproportionately affect
high-load players. Ensuring that players above a rolling-minutes threshold are
automatically rested on B2B nights — regardless of tier — would directly address this
interaction.

---

## Limitations

::: {.callout-important}
**Rare events**: ~0.29% rupture rate means any Linear Probability Model on this
outcome will have wide confidence intervals. This is a data constraint, not a
modelling failure. The NBA has 53 total ruptures over 35 years.

**Parallel trends**: the assumption holds well pre-2017 in the Low B2B panel.
The High B2B panel shows more noise — Model 2 results are directional only.

**Player tier proxy**: star status is approximated using season-level scoring
percentile rank. This captures the concept but is imperfect — a player who scores
highly in one season may not be an All-Star, and vice versa.

**Unobserved confounders**: off-season training loads, biomechanical factors, and
conditioning regimes are not captured in box score data.

**Future improvement**: a synthetic control approach would construct a weighted
counterfactual per player without relying on the parallel trends assumption.
:::

---

## The Bridge

We have established that load and B2B patterns are associated with the rupture spike
at the population level. But DiD answers *whether* and *why* — not *who* is next.

The same load features that explain the spike now become the input features of an
individual-level prediction engine. The causal analysis tells us **what matters**;
the predictive model tells us **who is at risk**.

→ [Bench the Instinct — Predictive Model](part2_predictive.qmd){.btn .btn-danger}

---

## References

Amin, N. H., Old, A. B., Tabb, L. P., Garg, R., Toossi, N., & Cerynik, D. L. (2013).
Performance outcomes after repair of complete Achilles tendon ruptures in National
Basketball Association players. *The American Journal of Sports Medicine, 41*(8),
1864–1868. https://doi.org/10.1177/0363546513490659

Farmer, S. (2013, April 14). Kobe Bryant has Achilles tendon surgery, out 6 to 9 months.
*Los Angeles Times*.

Fischer, B. (2023, May 17). Three Achilles injuries in one NBA postseason: What's going
on and how is it fixed? *Fox Sports*.

Goodwill, C. L. (2024, May 15). From 1990–2023, 45 NBA players tore their Achilles.
*Yahoo Sports*. https://sports.yahoo.com/article/1990-2023-45-nba-players-173643884.html

Hart, T. (2019, June 11). How Achilles injuries affected these nine NBA players.
*Sports Illustrated*.

Kuhn, M., & Silge, J. (2022). *Tidy modeling with R*. O'Reilly Media. https://www.tmwr.org

LaPrade, C. M., Chona, D. V., Cinque, M. E., Freehill, M. T., McAdams, T. R.,
Abrams, G. D., Sherman, S. L., & Safran, M. R. (2022). Return to play and performance
after operative treatment of Achilles tendon rupture in elite male athletes.
*British Journal of Sports Medicine, 56*(9), 515–520.
https://doi.org/10.1136/bjsports-2021-104835

Lewis, D. (2014, October 7). Kobe Bryant: Orthopaedic surgeon discusses star's Achilles
tendon, knee injuries. *Bleacher Report*.

Meadows, J., et al. (2024). Economic and performance analysis of NBA Achilles ruptures.
*Journal of Sports Economics* [forthcoming].

NBA.com. (2024, May 18). NBA Commissioner Adam Silver: League researching Achilles tears.
https://www.nba.com/news/nba-commissioner-adam-silver-league-researching-achilles-tears

Petway, A. J., Jordan, M. J., Epsley, S., & Anloague, P. A. (2022). Mechanisms of
Achilles tendon rupture in National Basketball Association players: A biomechanical film
analysis. *Journal of Applied Biomechanics, 38*(6), 398–403.
https://doi.org/10.1123/jab.2022-0088

Reynolds, T. (2025). NBA, Adam Silver address Achilles injury spike. *Associated Press*.
https://apnews.com/article/nba-adam-silver-achilles-injuries-b81beee74cee07ac2d644162be835d7c

Silver, A. (2025). Adam Silver says league taking serious look at Achilles tears. *ESPN*.
https://www.espn.com.sg/nba/story/_/id/45585272

Sportico. (2022). NBA schedule reduces back-to-backs, travel to improve game quality.
*Sportico*. https://www.sportico.com/leagues/basketball/2022/nba-schedule-2023-rest-travel-1234691876/

Siu, R., Lee, J., & Chan, J. (2020). Prognosis of elite basketball players after an
Achilles tendon rupture. *Science Progress, 103*(3).
https://doi.org/10.1177/0036850420936180

Trofa, D. P., Miller, J. C., Jang, E. S., Woode, D. R., Greisberg, J. K., &
Vosseller, J. T. (2017). Professional athletes' return to play and performance after
operative repair of an Achilles tendon rupture. *The American Journal of Sports
Medicine, 45*(12), 2864–2871. https://doi.org/10.1177/0363546517713001

Windhorst, B. (2017). NBA outlines schedule changes in memo to teams. *ESPN*.
https://www.espn.com/nba/story/_/id/20286965/nba-issues-memo-detailing-h
