---
title: "Predicting Load-Related Game Unavailability in the NBA"
subtitle: "A Causal and Predictive Analysis of Achilles Injury Burden"
date: "2026 April"
author:
  - name: Keziah Vickraman
    affiliations:
      - name: "SMU Academy"
abstract: |
  Achilles tendon ruptures represent some of the most severe and career-disrupting
  injuries in professional basketball. The 2024–25 NBA season recorded eight confirmed
  ruptures — nearly six times the historical per-season average of 1.36. This study
  investigates the role of player workload in driving this spike using two complementary
  analytical frameworks. First, a Difference-in-Differences (DiD) design is used to
  estimate whether the shift toward higher rolling-minutes loads and more frequent
  back-to-back scheduling has causally contributed to elevated rupture rates,
  particularly since 2016 when systematic load management emerged as a league concern.
  Second, an XGBoost regression model is trained on NBA box score data (2018–2025)
  to predict individual player game unavailability — operationalised as the percentage
  of team games missed in a season — using the same load features that explain the
  population-level trend. The causal analysis finds directionally positive DiD
  interaction coefficients for high-load × post-2016 player-seasons, with amplification
  in back-to-back contexts. The predictive model achieves meaningful discrimination on
  a held-out temporal test set (2020–2025), with prior missed games, rolling minutes,
  and career tenure emerging as the strongest predictors. Together, these analyses
  provide a data-driven foundation for proactive load management decisions in elite
  basketball.
keywords:
  - NBA
  - Achilles tendon
  - load management
  - difference-in-differences
  - XGBoost regression
  - sports analytics
  - injury prediction
format:
  html:
    toc: true
    number-sections: true
    bibliography: references.bib
    csl: apa.csl
  pdf:
    toc: true
    number-sections: true
    bibliography: references.bib
    csl: apa.csl
    documentclass: article
    geometry: margin=1in
    fontsize: 12pt
    linestretch: 1.5
    colorlinks: true
    link-citations: true
    keep-tex: false
    include-in-header:
      text: |
        \usepackage{lscape}
---
## Executive Summary

### The Observation

The 2024–25 NBA season recorded **8 confirmed Achilles tendon ruptures** — nearly
six times the historical average of 1.36 per season. The names are not fringe
players. Jayson Tatum. Tyrese Haliburton. Damian Lillard. Dejounte Murray.
This is not a coincidence. This is a signal.

NBA Commissioner Adam Silver convened an expert panel and pointed directly at
workload intensity and scheduling density as the primary suspects. This project
is the data response to that question.

---

### Part 1 — Did Load Kill the Achilles?

**Observation:** Achilles rupture rates have risen disproportionately among
high-load players since 2016 — the year the NBA began formally discussing
load management protocols.

**Empirical Evidence:** A Difference-in-Differences (DiD) analysis comparing
high-load vs low-load player-seasons before and after 2016 finds a
**positive interaction coefficient** across all three model specifications.
The effect is amplified for players with heavy back-to-back scheduling exposure.
Base rupture rate: ~0.29% — rare, but the directional signal is consistent and
theoretically grounded.

**Interpretation:** Load management protected the stars. The 75th percentile
of player minutes plateaued post-2016 while median player loads increased —
widening the gap. High-usage rotation players outside formal rest protocols
are the most structurally exposed group.

**Recommendation:** Load management protocols must extend beyond max-contract
stars to the full roster. Back-to-back scheduling for high-load players should
be systematically avoided. Inconsistent load management — protecting some
while increasing others' exposure — may be worse than no protocol at all.

---

### Part 2 — Who's Next?

**Observation:** The same load features that explain the population-level
spike can be used to flag individual players before injury occurs. The model
correctly identified Zach LaVine, LaMelo Ball, Joel Embiid, Gordon Hayward,
and OG Anunoby as top at-risk players for the 2024–25 season.

**Empirical Evidence:** An XGBoost regression model trained on 2018–2021 NBA
box score data and tested on 2022–2025 achieves **R² = 0.747, RMSE = 0.133,
MAE = 0.100** on the held-out test set. The three strongest predictors are
`prior_missed_pct` (last season's absence rate), `avg_rolling_minutes_3`
(lagged 3-game rolling average), and `career_season` (years in the league —
risk peaks at seasons 4–11).

**Interpretation:** Luka Doncic was flagged as Moderate Risk based on his
2024 load profile — 35.8 avg minutes, 8th career season, prior hamstring
history. He subsequently missed ~22% of the 2025–26 regular season with a
Grade 2 hamstring strain. The model does not predict injury type or timing.
It predicts load-related game unavailability — and the signal is real.

**Recommendation:** Teams should run this model before each season on their
full roster, not just stars. A player predicted to miss 30%+ of games is a
roster construction problem, not just a medical one. The interactive dashboard
allows coaches and GMs to simulate load reduction scenarios in real time —
shifting the conversation from reactive treatment to proactive prevention.

---

::: {.callout-important}
## The Bottom Line
At an average cost of **$4M per Achilles rupture** — and up to **$70M for a
max-contract player** — the return on a data-driven load management programme
is clear. The gap between intention and action in injury prevention is not a
data problem. It is a tools problem. This project is one step toward closing it.
:::
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  dev     = "ragg_png"
)
source("00_Data_Setup.R")
load("Part1_Causal_Results.RData")
load("Part2_Predictive_Results.RData")
```

---

# Introduction

Achilles tendon ruptures are among the most catastrophic musculoskeletal injuries
sustained by professional athletes, particularly in high-impact sports such as
basketball. In the National Basketball Association (NBA), where explosive movements,
rapid directional changes, and consistently high workloads are the norm, a ruptured
Achilles tendon can significantly disrupt or prematurely end a player's career
[@Amin2013; @Trofa2017].

Although less frequent than other lower extremity injuries, Achilles ruptures carry
disproportionately severe consequences. High-profile cases — Kobe Bryant (2013),
Kevin Durant (2019), and more recently Jayson Tatum and Tyrese Haliburton (2025)
— have drawn sustained public attention to the injury's physical, psychological,
and financial toll. The average recovery timeline ranges from 9 to 12 months, and
studies consistently show lasting declines in performance metrics even for players
who return to competition [@Trofa2017; @Siu2020].

The 2024–25 NBA season has brought renewed urgency to this issue. Eight confirmed
Achilles ruptures have been recorded to date — nearly six times the historical
per-season average of 1.36 cases documented between 1990 and 2023 [@Goodwill2024].
NBA Commissioner Adam Silver acknowledged the league is "taking a serious look" at
Achilles tears and convened an expert panel, citing workload intensity and scheduling
density as primary investigative targets [@Silver2025; @Reynolds2025].

This study responds to that concern with data. We investigate two related questions:

1. **Causal**: Did the NBA's shift toward higher player loads and more frequent
   back-to-back scheduling *cause* an elevated Achilles rupture rate, particularly
   since 2016?

2. **Predictive**: Using the same load features, which individual players face the
   highest risk of missing significant game time in the coming season?

The two analyses are connected by a deliberate design choice: the features that
explain the population-level spike in Part 1 are the input features of the
individual-level prediction engine in Part 2. Load is the analytical bridge.

# Literature Review

Research on NBA Achilles injuries has primarily focused on post-injury recovery and
return-to-play timelines. @Amin2013 documented that most players return to competition
following surgical repair but experience lasting declines in performance metrics.
@Trofa2017 confirmed this pattern in a systematic review, finding that performance
decreases were particularly pronounced among older and high-usage players.

Biomechanical analyses have begun to identify pre-injury loading patterns. @Petway2022
conducted film-based analysis of NBA gameplay footage and identified specific landing
mechanics commonly preceding ruptures, suggesting that such injuries may not be entirely
random. @LaPrade2022 emphasised the role of cumulative fatigue and overuse —
especially during congested scheduling periods — as likely contributors to tendon failure.

The case for proactive prediction has been articulated by @Siu2020, who argued that
early identification of at-risk players can support preventative strategies including
load management, individualised rest protocols, and targeted prehabilitation.
@Maffulli2011 estimated that up to 70% of Achilles ruptures and tendinopathies are
preventable with appropriate intervention — a figure that frames the economic and
competitive stakes of the present analysis.

# Data

## Sources

Three data sources are used across this study (@tbl-data):

```{r tbl-data}
#| tbl-cap: "Data sources"
tibble::tribble(
  ~Source,                    ~Coverage,         ~`Used For`,
  "hoopR (NBA box scores)",   "2018–2025",       "Game-level features, pct_games_missed (Part 2)",
  "Pro Sports Transactions",  "1990–2023",       "Confirmed Achilles ruptures (Part 1)",
  "ESPN injury log (scraped)","2024–25 season",  "Recent ruptures + injury flags",
  "Manual curation",          "2024–25 season",  "8 confirmed 2024–25 Achilles ruptures"
) |>
  knitr::kable(align = c("l","l","l"))
```
## NBA Box Scores

Regular season game-level data were retrieved for seasons 2002–2025 using the
`hoopR` R package [@saiemgilani2023]. Playoff data were excluded from the primary
analysis to maintain scheduling consistency. The resulting dataset contains
approximately 758606 player-game observations across 2458 unique players.

## Achilles Rupture Records

Confirmed rupture cases were extracted from the Kaggle NBA Player Injury Stats
dataset (sourced from ProSports Transactions) for the period 1990–2023.
Cases were identified using keyword detection on injury descriptions
(`rupture`, `tear`, `torn`, `surgery`, `repair`, `reconstruction`), restricting
to entries explicitly mentioning the Achilles tendon. Only the first recorded
rupture per player was retained to avoid double-counting multi-season absences.

The 2024–25 season cases (n = 8) were manually curated from verified reports
in ESPN and Yahoo Sports due to the recency of these events.

## Target Variable Construction

The predictive target, `pct_games_missed`, is defined as:

$$\text{pct\_games\_missed}_{i,t} = 1 - \frac{\text{games appeared}_{i,t}}{\text{team games played}_{i,t}}$$

where $i$ indexes players and $t$ indexes seasons. The denominator uses the
player's primary team's game count for that season, accommodating mid-season
trades. Values are bounded $[0, 1]$; 0 indicates the player appeared in every
available game.

# Methodology

## Part 1: Difference-in-Differences

### Design

A Difference-in-Differences estimator compares the change in Achilles rupture
rates between high-load and low-load player-seasons, before and after 2016.
The treatment indicator, $D_i$, is assigned to player-seasons in the top
quartile of rolling 3-game average minutes, computed within each season to
account for era-level differences in player usage.

The primary estimating equation is:

$$\text{Rupture}_{i,t} = \alpha + \beta_1 D_i + \beta_2 \text{Post}_t + \beta_3 (D_i \times \text{Post}_t) + \epsilon_{i,t}$$

The DiD estimator $\hat{\beta}_3$ captures the differential change in rupture
probability for high-load players after 2016, relative to the low-load control
group. A Linear Probability Model (LPM) is used throughout for coefficient
interpretability; all estimates are reported as percentage-point changes in
rupture probability.

### Post-Period Cutoff

The 2016–17 season marks the emergence of organised load management as a
league-wide practice following CBA renegotiation. This provides a theory-grounded,
non-data-driven cutoff — a necessary condition for DiD identification.

### B2B Interaction

A triple-interaction model extends the baseline specification:

$$\text{Rupture}_{i,t} = \alpha + \beta (D_i \times \text{Post}_t \times \text{B2B}_{i,t}) + \text{controls} + \epsilon_{i,t}$$

where $\text{B2B}_{i,t}$ indicates above-median back-to-back game exposure.
This tests whether the load effect is amplified for players with heavy
scheduling density — directly addressing Commissioner Silver's stated concern.

## Part 2: XGBoost Regression

### Architecture

A direct XGBoost regression model predicts `pct_games_missed` for each
player-season. A hurdle model was initially considered given the zero-inflated
structure of the outcome, but diagnostic analysis revealed that 87–94% of
player-seasons involve at least some missed games in the reliable data window
(2018–2025), making the binary Stage A uninformative. The single-stage regression
approach is more appropriate given this distribution.

The model was tuned using a space-filling grid of 15 hyperparameter combinations
across tree depth, learning rate, loss reduction, sample proportion, and mtry,
evaluated via time-aware cross-validation using `sliding_period()` with 1-year
lookback windows.

### Survivorship Bias Correction

Partial dependence plot (PDP) analysis revealed that `total_minutes`,
`back_to_back_games`, and `avg_days_rest` all showed negative relationships
with `pct_games_missed` — the opposite of the load management hypothesis.
This reflects survivorship bias: players who suffered early-season injuries
accumulated fewer total minutes and fewer B2B games precisely because they
were unavailable. These three features were removed from the final model.

`avg_rolling_minutes_3` (lagged 3-game rolling average) was retained as the
primary load predictor — it shows a clean positive monotonic relationship with
games missed because it measures load *before* the outcome period, escaping
the survivorship problem.

### Train/Test Split

Training data span 2018–2021. The test set spans 2022–2025. The 2018 cutoff
reflects the earliest season where hoopR box score coverage is sufficiently
complete for reliable `pct_games_missed` computation — earlier seasons show
systematically undercounted player appearances.

# Results

## Part 1: DiD Results

```{r fig-parallel, fig.height=5, fig.cap="Parallel trends check. Pre-2016 trends are broadly parallel between load groups, particularly in the low B2B panel."}
did_parallel_trends_VIZ_plain

```

@fig-parallel shows the mean Achilles rupture rate by season, load group, and
B2B exposure. Pre-2016 trends are broadly parallel in the low-B2B panel,
supporting the parallel trends assumption for that subsample. The high-B2B
panel shows more pre-treatment noise.


```{r did-results, results='asis'}
if (knitr::is_html_output()) {
  export_summs(
    model_did_1, model_did_2, model_did_3,
    model.names  = c("Model 1\nLoad × Post",
                     "Model 2\nLoad × Post × B2B",
                     "Model 3\nContinuous"),
    error_format = "CIs: [{conf.low}, {conf.high}]",
    digits       = 3
  )
}
```


The DiD interaction coefficient (`High Load × Post-2016`) is positive across
all three specifications (Table 2), consistent with the hypothesis that
high-load player-seasons carry elevated rupture risk in the post-2016 period.
The triple interaction model (Model 2) indicates further amplification when
high load is combined with heavy back-to-back exposure.

Confidence intervals are wide — consistent with the rarity of the outcome
(~0.29% base rate). Results should be interpreted directionally.

## Part 2: Predictive Results

```{r fig-roc, fig.height=4.5}
feature_imp_VIZ_plain
```

The feature importance chart shows that `prior_missed_pct` (last season's game unavailability)
is the strongest single predictor of future unavailability, followed by load
features — consistent with both the DiD findings and the clinical literature on
injury recurrence. Career tenure (`career_season`) also ranks highly, reflecting
the cumulative wear on tendon tissue over long careers.

```{r fig-risk, fig.height=6}
risk_table_VIZ_plain
```

# Discussion

This study makes two contributions. First, the DiD analysis provides causal
evidence — albeit directional rather than definitive, given data constraints —
that the NBA's shift toward higher load environments is associated with elevated
Achilles rupture rates. The amplification in back-to-back contexts directly
corroborates the concerns raised by Commissioner Silver and the NBA's expert panel.

Second, the XGBoost regression model translates this population-level finding into
individual-level actionability. The what-if tool allows medical staff to
simulate the effect of load reduction on a specific player's predicted risk
before the season begins — shifting the intervention window from reactive
(post-rupture rehabilitation) to proactive (pre-season load planning).

The convergence of findings across both analytical frameworks — where the same
features explain the population trend and drive the individual predictions —
strengthens confidence in the load management hypothesis. This is not a
coincidence of model selection; it reflects a genuine data signal.

# Conclusion

The 2024–25 NBA Achilles rupture spike is not simply bad luck. Load patterns —
measurable, monitorable, and modifiable — are directionally associated with
elevated injury risk at both the population and individual levels. A data-driven
load management programme, informed by tools like those developed in this study,
could prevent ruptures before they occur.

At an estimated cost of $4M per incident on average — and up to $70M for
max-contract players [@Meadows2024] — the return on investment is clear.
The gap between intention and action in injury prevention is not a data problem.
It is a tools problem. This project is one step toward closing it.

# Acknowledgements

This project was completed as part of the capstone requirement at **SMU Academy**
under the Certified Data Analyst (R) Specialist course. 
I would like to thank **Professor Roh** for his guidance, feedback, and supervision throughout the course of this project.

I also wish to acknowledge my teammates **Bibi Haneeza Said** and **Eugene Tan**
for their contributions during the initial phases of the capstone, particularly
in data collection and exploratory analysis.

The following datasets and tools were used in this study:

- **hoopR** (Gilani, 2023) — NBA player box score data (2002–2025), accessed via
  the `hoopR` R package from the Sports Data Verse
- **NBA Player Injury Stats (1951–2023)** — sourced from
  [Kaggle](https://www.kaggle.com), originally compiled from ProSports Transactions
- **ESPN Injury Log** — web-scraped from
  [espn.com.sg/nba/injuries](https://www.espn.com.sg/nba/injuries) for the
  2024–25 season
- **R** and the `tidymodels`, `xgboost`, `ggplot2`, and `shiny` ecosystems
  were used for all data processing, modelling, and deployment

# References

::: {#refs}
:::
